---
title: NCAA Tournament Simulation
output: html_document
runtime: shiny

---

```{r Install/Load Packages, include=FALSE}

require(tidyverse)
require(shiny)
require(knitr)
library(rsconnect)

```

```{r Load Kenpom Data, include=FALSE}

ratings <- read.csv("summary20.csv")

```

```{r Define Functions, include=FALSE}

game_predict <- function(ratings, high_seed, low_seed){
  
# Calculate Pythagorean Exp Win % for all teams
ratings <- ratings %>% mutate(Pyth = AdjOE ^ 11.5 / (AdjOE ^ 11.5 + AdjDE ^ 11.5))

# NCAA D1 National Averages
avgT <- mean(ratings$AdjTempo) 
avgE <- mean(ratings$AdjOE)

# High and low seed parameters
tempo_high <- subset(ratings, TeamName == high_seed) %>% select(AdjTempo)
tempo_low <- subset(ratings, TeamName == low_seed) %>% select(AdjTempo)
adjO_high <- subset(ratings, TeamName == high_seed) %>% select(AdjOE)
adjO_low <- subset(ratings, TeamName == low_seed) %>% select(AdjOE)
adjD_high <- subset(ratings, TeamName == high_seed) %>% select(AdjDE)
adjD_low <- subset(ratings, TeamName == low_seed) %>% select(AdjDE)
AdjEM_high <- subset(ratings, TeamName == high_seed) %>% select(AdjEM)
AdjEM_low <-  subset(ratings, TeamName == low_seed) %>% select(AdjEM)
pyth_high <- subset(ratings, TeamName == high_seed) %>% select(Pyth)
pyth_low <- subset(ratings, TeamName == low_seed) %>% select(Pyth)

# Matchup-specific parameters
game_tempo <- tempo_high + tempo_low - avgT
ORtg_high <-  adjO_high + adjD_low - avgE
ORtg_low <-  adjO_low + adjD_high - avgE

# Prediction & Odds
prob_high <- (pyth_high - pyth_high * pyth_low) / (pyth_high + pyth_low - 2 * pyth_high *
              pyth_low)
prob_low <- (pyth_low - pyth_low * pyth_high) / (pyth_low + pyth_high - 2 * pyth_low *
              pyth_high)

win_pct <<- round(max(prob_high, prob_low) * 100, digits = 0)

pts_high <- ORtg_high * game_tempo / 100
pts_low <- ORtg_low * game_tempo / 100

pts_high_round <- round(pts_high, digits = 0)
pts_low_round <- round(pts_low , digits = 0)

winner <<- if(pts_high > pts_low) high_seed else low_seed
loser  <<- if(pts_high > pts_low) low_seed else high_seed

win_score <- if (pts_high_round == pts_low_round && pts_high > pts_low) {
                 pts_high_round + 1 } else 
                                      if (pts_high_round == pts_low_round && pts_low > pts_high) {
                                          pts_high_round + 1} else
                                                              max(pts_high_round, pts_low_round)
lose_score <- min(pts_high_round, pts_low_round)
  
paste("kenpom.com prediction: ",paste(paste(winner, paste(win_score, lose_score, sep = "-"), sep = " ")," (",win_pct,"%",")",sep = ""), sep = "")

}

## Write function to predict a single instance
game_sim <- function(ratings, high_seed, low_seed){

game_predict(ratings, high_seed, low_seed)
  
rand <- floor(runif(1, min=0, max=101))

# Generate random integer between 1-100
# The favorite wins if number is <= predicted win probability
msg <- if(rand <= win_pct) winner else loser

paste("The basketball gods chose", msg, "- they advance to the next round!", sep = " ")
}

```
\
\
\
\
\
\
\
\
\
\
\
\
\
\
\


This was a project I put together when I learned that the 2020 NCAA Tournament was cancelled due to the COVID-19 pandemic. As the organizer of my company bracket pool, I developed this app as a way to simulate the results of the tournament and give some levity to some of the grave news we were getting every day. I encourage you to play around with it.

The app shows, for a hypothetical matchup of two teams, the predicted score and win probability for the favored team.
Predictions are based on the ratings compiled by Ken Pomeroy at kenpom.com. Clicking the 'Simulate' button below will simulate one game of the selected matchup, weighted according to the odds derived from the kenpom.com ratings.

## Select Matchup


```{r Predictions and Simulation, echo = FALSE}

shinyApp(
    shinyUI(
      fluidPage(
        inputPanel(
          selectInput(
              'high_seed', label = 'High Seed',
               choices = ratings$TeamName, selected = "Villanova"),
          selectInput(
              'low_seed', label = 'Low Seed',
              choices = ratings$TeamName, selected = "Dayton")
        ),
      tableOutput("result"),
      actionButton("simulate", "Simulate"),
       verbatimTextOutput("sim")
      )
    ),
  shinyServer(function(input, output) {
    output$result <- renderText({
      result <- game_predict(ratings, input$high_seed, input$low_seed)})
    
    observeEvent(input$simulate, {
      output$sim <- renderText(game_sim(ratings, input$high_seed, input$low_seed))})})
)

```
 Data current as of March 11, 2020.
